---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# weightedcontrasts

<!-- badges: start -->
<!-- badges: end -->

Provides the function `contr.poly.weighted` to apply orthogonal polynomial
contrasts to unbalanced data. 
  
## Installation

You can install the development version:

``` r
remotes::install_github("elbersb/weightedcontrasts")
```

## Example

```{r example}
library("weightedcontrasts")
## basic example code
```

## References

Elbers, Benjamin. 2020. [_Orthogonal Polynomial Contrasts and Applications to Age-Period-Cohort Models_](https://osf.io/preprints/socarxiv/xrbgv/), Working Paper.

Fosse, Ethan and Christopher Winship. 2019. [_Analyzing Age-Period-Cohort Data: A Review and Critique_](https://www.annualreviews.org/doi/abs/10.1146/annurev-soc-073018-022616), Annual Review of Sociology 45:467–92.

Holford, Theodore R.. 1983. [_The Estimation of Age, Period and Cohort Effects for Vital Rates_](https://www.jstor.org/stable/2531004), Biometrics 39(2): 311-324.
