---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# weightedcontrasts

<!-- badges: start -->
[![CRAN Version](https://www.r-pkg.org/badges/version/weightedcontrasts)](https://CRAN.R-project.org/package=weightedcontrasts)
[![R-CMD-check](https://github.com/elbersb/weightedcontrasts/workflows/R-CMD-check/badge.svg)](https://github.com/elbersb/weightedcontrasts/actions)
[![Codecov test coverage](https://codecov.io/gh/elbersb/weightedcontrasts/branch/master/graph/badge.svg)](https://codecov.io/gh/elbersb/weightedcontrasts?branch=master)
<!-- badges: end -->

Provides the function `contr.poly.weighted` to apply orthogonal polynomial
contrasts to unbalanced data. The function is general, but the examples are
specific to age-period-cohort models. Currently, the package contains the following:

- The function `contr.poly.weighted`
- The dataset `prostate` from Holford 1983 (Table 2)
- A vignette that reanalyzes [the example given in Holford 1983](https://htmlpreview.github.io/?https://github.com/elbersb/weightedcontrasts/blob/master/doc/fosse_winship2019.html)
- A vignette that reanalyzes [the example given in Fosse and Winship 2019](https://htmlpreview.github.io/?https://github.com/elbersb/weightedcontrasts/blob/master/doc/holford1983.html)

## Installation

You can install the development version:

``` r
remotes::install_github("elbersb/weightedcontrasts")
```

## Example

This example shows how `contr.poly.weighted` provides a contrast matrix that will lead
to an orthogonal design matrix, even if the data are unbalanced. We assume a situation in which
an evenly-spaced predictor, such as age groups (e.g., 20-24, 25-29, 30-34, etc.).

```{r example}
library("weightedcontrasts")

# first, the balanced case
group <- factor(c(20, 20, 25, 25, 30, 30))
contrasts(group) <- contr.poly
X <- model.matrix(~ group)
zapsmall(crossprod(X)) # correct result

# but in the unbalanced case, contr.poly fails
group <- factor(c(20, 20, 25, 25, 30))
contrasts(group) <- contr.poly
X <- model.matrix(~ group)
zapsmall(crossprod(X)) # wrong result!

# this is where contr.poly.weighted comes in
# (width specifies the width of the group intervals)
group <- factor(c(20, 20, 25, 25, 30))
contrasts(group) <- contr.poly.weighted(group, width = 5)
X <- model.matrix(~ group)
zapsmall(crossprod(X)) # correct result
```

## References

Elbers, Benjamin. 2020. [_Orthogonal Polynomial Contrasts and Applications to Age-Period-Cohort Models_](https://osf.io/preprints/socarxiv/xrbgv/), Working Paper.

Fosse, Ethan and Christopher Winship. 2019. [_Analyzing Age-Period-Cohort Data: A Review and Critique_](https://www.annualreviews.org/doi/abs/10.1146/annurev-soc-073018-022616), Annual Review of Sociology 45:467–92.

Holford, Theodore R.. 1983. [_The Estimation of Age, Period and Cohort Effects for Vital Rates_](https://www.jstor.org/stable/2531004), Biometrics 39(2): 311-324.
