---
title: "A Reanalysis of Fosse and Winship (2019)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A Reanalysis of Fosse and Winship (2019)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

This is a reanalysis of an example provided in Ethan Fosse and Christopher Winship, 2019, [_Analyzing Age-Period-Cohort Data: A Review and Critique_](https://www.annualreviews.org/doi/abs/10.1146/annurev-soc-073018-022616), Annual Review of Sociology 45:467–92. The  example is based on General Social Survey (GSS) data for N=23,825 respondents. The outcome is the number of words correct on a vocabulary quiz of ten items (min: 0, max: 10, sd: 2.14). Age and period are grouped into five-year intervals.

For details on these results, see [this working paper](https://osf.io/preprints/socarxiv/xrbgv/).

## Load packages and data

```{r message=FALSE, warning=FALSE}
library("dplyr")
library("ggplot2")
library("texreg")
library("weightedcontrasts")

data("wordsum") # from apcR package (https://scholar.harvard.edu/apc/software-0)
d <- as_tibble(wordsum)
```

## Define helper functions

The first helper function extracts the non-linear effects from the model, while the second
function computes a total (linear and non-linear) effect. The third function simply
constructs a data frame from given age, period, and cohort effects, and is used for plotting.

```{r}
get_deviations_w_intercept <- function(model, set, contrasts) {
    extract_coefs <- coef(model)[grepl(set, names(coef(model)))]
    extract_coefs <- extract_coefs[2:length(extract_coefs)]
    extract_coefs[is.na(extract_coefs)] <- 0
    deviations <- contrasts[, 2:(1 + length(extract_coefs))] %*% extract_coefs
    coef(model)[1] + deviations[, 1]
}

get_total_effect <- function(contrasts, linear_coef, nonlinear_coefs) {
    linear_contrast <- contrasts[, 1]
    nonlinear_contrasts <- contrasts[, 2:ncol(contrasts)]
    linear_contrast * linear_coef + nonlinear_contrasts %*% nonlinear_coefs
}

get_apc_df <- function(age, period, cohort) {
    tibble(effect = c(age, period, cohort),
        year = c(names(age), names(period), names(cohort)),
        apc = c(rep("Age", length(age)),
            rep("Period", length(period)),
            rep("Cohort", length(cohort))))
}
```

## Replicating Fosse & Winship's original model

We first replicate Fosse and Winship's original model. The function `apcR::create.poly` uses
unweighted orthogonal polynomials. We then use one of the helper functions defined above to extract
the identified non-linear effects.

```{r}
data_fw <- as_tibble(d)

contrasts(data_fw$a) <- apcR::create.poly
contrasts(data_fw$p) <- apcR::create.poly
contrasts(data_fw$c) <- apcR::create.poly
# zero out the period linear effect
contrasts(data_fw$p)[, 1] <- 0

fw <- lm(wordsum ~ a + c + p, data = data_fw)

# get deviations from linear trend and age/cohort linear effects
# (these are the true linear effects when period linear effect = 0)
fw_age <- get_deviations_w_intercept(fw, "^a", contrasts(data_fw$a))
fw_period <- get_deviations_w_intercept(fw, "^p", contrasts(data_fw$p))
fw_cohort <- get_deviations_w_intercept(fw, "^c", contrasts(data_fw$c))
fw_theta1 <- coef(fw)["a.L"]
fw_theta2 <- coef(fw)["c.L"]
```

## Using weighted orthogonal contrasts

This code is very similar, but uses the function `contr.poly.weighted` from the [`weightedcontrasts`](https://github.com/elbersb/weightedcontrasts) package.
We also specify the width of the age groups (5-year intervals in this case), to obtain
an interpretable linear term.

```{r}
data_corrected <- as_tibble(d)

contrasts(data_corrected$a) <- contr.poly.weighted(data_corrected$a, width = 5)
contrasts(data_corrected$p) <- contr.poly.weighted(data_corrected$p, width = 5)
contrasts(data_corrected$c) <- contr.poly.weighted(data_corrected$c, width = 5)
# zero out the period linear effect
contrasts(data_corrected$p)[, 1] <- 0

corrected <- lm(wordsum ~ a + c + p, data = data_corrected)

# get deviations from linear trend and age/cohort linear effects
# (these are the true linear effects when period linear effect = 0)
corrected_age <- get_deviations_w_intercept(corrected, "^a", contrasts(data_corrected$a))
corrected_period <- get_deviations_w_intercept(corrected, "^p", contrasts(data_corrected$p))
corrected_cohort <- get_deviations_w_intercept(corrected, "^c", contrasts(data_corrected$c))
corrected_theta1 <- coef(corrected)["a.L"]
corrected_theta2 <- coef(corrected)["c.L"]
```

## Comparison

The model estimates can be seen in this table. The differences arise from the use
of unweighted vs. weighted orthogonal polynomials, and from the fact that the linear terms
are interpretable as the change in one year in the corrected model.

```{r results="asis"}
htmlreg(list(fw, corrected), single.row = TRUE)
```

We can also plot the non-linear effects:

```{r}
to_plot <- bind_rows(
    get_apc_df(corrected_age, corrected_period, corrected_cohort) %>%
        mutate(which = "This paper"),
    get_apc_df(fw_age, fw_period, fw_cohort) %>%
        mutate(which = "Fosse & Winship 2019")) %>%
    mutate(which = factor(which, levels = c("This paper", "Fosse & Winship 2019")),
        apc = factor(apc, levels = c("Age", "Period", "Cohort")),
        year = as.numeric(year))

# comparison of non-linear effects
p <- to_plot %>%
    ggplot(aes(x = year, y = effect, group = which, linetype = which)) +
    facet_wrap("apc", scales = "free_x") +
    geom_point() + geom_line() +
    labs(y = "Non-linear effect", group = "", linetype = "") +
    theme_bw() +
    theme(legend.position = "bottom", axis.title.x = element_blank())
p
```

```{r, include = FALSE}
# ggsave("fosse_winship2019.pdf", p, height = 4, width = 7)
```

## 2D-APC Plot

Fosse and Winship introduced the idea of a 2D-APC plot in their work. These plots are reproduced
here for the two sets of results.

```{r}
two2d_plot <- function(theta1, theta2) {
    sum_thetas = abs(theta2 - theta1)
    limits = c(-sum_thetas, sum_thetas)

    ggplot() +
        geom_abline(aes(intercept = theta1, slope = -1),
            size = 1, color = "#1080BA") +
        geom_hline(yintercept = 0, linetype = 2, color = "#3C4650") +
        geom_hline(yintercept = -(theta2 - theta1), linetype = 2, color = "#3C4650") +
        geom_vline(xintercept = 0, linetype = 2, color = "#3C4650") +
        scale_y_continuous(expression(alpha),
            limits = limits * 1.2,
            sec.axis = sec_axis(~.+(theta2 - theta1), name = expression(gamma))) +
        scale_x_continuous(expression(pi),
            limits = limits * 1.1,
            sec.axis = dup_axis()) +
        theme_bw() +
        theme(text = element_text(size = 10),
            axis.title.y.left = element_text(angle = 0, vjust = 0.5),
            axis.title.y.right = element_text(angle = 0, vjust = 0.5))
}

two2d_plot(fw_theta1, fw_theta2)
two2d_plot(corrected_theta1, corrected_theta2)
```

```{r, include = FALSE}
# ggsave("fosse_winship2019_2dapc.pdf", p, height = 3, width = 4)
```

## Obtaining smoothed estimates

Orthogonal polynomials can be used as simple smoothers, by simply dropping higher-order
polynomial terms. The figure below shows cohort non-linear effects when modeled using different degrees,
from low to high. The most complex model will use polynomials of degree 19.

```{r fig.width = 8}
devs <- list()
for (i in seq(ncol(contrasts(data_corrected$c)), 3)) {
    contrasts(data_corrected$c)[, i] <- 0
    reduced <- lm(wordsum ~ a + c + p, data = data_corrected)
    devs_i <- get_deviations_w_intercept(reduced, "^c", contrasts(data_corrected$c))
    devs[[i]] <- tibble(degree = i - 1, year = names(devs_i), deviations = devs_i)
}

cohort_by_degree <- bind_rows(devs) %>%
    bind_rows(tibble(degree = ncol(contrasts(data_corrected$c)),
        year = names(corrected_cohort),
        deviations = corrected_cohort)) %>%
    mutate(year = as.numeric(year))

p <- cohort_by_degree %>%
    ggplot(aes(x = as.numeric(year), y = deviations)) +
    facet_wrap("degree") +
    geom_point() + geom_line() +
    theme_bw()
p
```

## Monotonicity constraint for cohort

Smoothed estimates are desirable when monotonicity constraints are specified. We choose
here, somewhat arbitrarily, a polynomial of degree 5, and then specify a monotonicity constraint
for the total cohort effect. The dashed line shows the minimum total cohort effect under the assumption
that it is monotonically increasing.

```{r}
choose_degree <- 5
non_lins <- filter(cohort_by_degree, degree == choose_degree)$deviations
min_slope <- -min(diff(non_lins)) / 5
total_cohort <- non_lins + contrasts(data_corrected$c)[, 1] * min_slope

cohort_df <- filter(cohort_by_degree, degree == choose_degree) %>%
    transmute(effect = deviations,
        year,
        which = paste0("Non-linear only, poly. of degree ", degree)) %>%
    bind_rows(tibble(effect = total_cohort,
        year = as.numeric(names(total_cohort)),
        which = "Under monotonicity constraint"))

p <- cohort_df %>%
    ggplot(aes(x = year, y = effect, linetype = which)) +
    geom_point() + geom_line() +
    theme_bw() +
    theme(text = element_text(size = 12), legend.position = "bottom") +
    labs(x = "Cohort", y = "Effect", linetype = "")
p
```

```{r, include = FALSE}
# ggsave("fosse_winship2019_monotonicity.pdf", p, height = 4, width = 6)
```
