---
title: "A Reanalysis of Fosse and Winship (2019)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A Reanalysis of Fosse and Winship (2019)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This is a reanalysis of an example provided in Ethan Fosse and Christopher Winship, 2019, [_Analyzing Age-Period-Cohort Data: A Review and Critique_](https://www.annualreviews.org/doi/abs/10.1146/annurev-soc-073018-022616), Annual Review of Sociology 45:467–92. The  example is based on General Social Survey (GSS) data for N=23,825 respondents. The outcome is the number of words correct on a vocabulary quiz of ten items (min: 0, max: 10, sd: 2.14). Age and period are grouped into five-year intervals.

## Load packages and data

```{r message=FALSE, warning=FALSE}
library("dplyr")
library("ggplot2")
library("texreg")
library("weightedcontrasts")

library("apcR") # required for dataset, get here: https://scholar.harvard.edu/apc/software-0
data("wordsum")
d <- as_tibble(wordsum)
```

## Define helper functions

```{r}
get_deviations_w_intercept <- function(model, set, contrasts) {
    extract_coefs <- coef(model)[grepl(set, names(coef(model)))]
    extract_coefs <- extract_coefs[2:length(extract_coefs)]
    extract_coefs[is.na(extract_coefs)] <- 0
    deviations <- contrasts[, 2:(1 + length(extract_coefs))] %*% extract_coefs
    coef(model)[1] + deviations[, 1]
}

get_total_effect <- function(contrasts, linear_coef, nonlinear_coefs) {
    linear_contrast <- contrasts[, 1]
    nonlinear_contrasts <- contrasts[, 2:ncol(contrasts)]
    linear_contrast * linear_coef + nonlinear_contrasts %*% nonlinear_coefs
}
```

## Replicating Fosse & Winship's original model

```{r}
data_fw <- as_tibble(d)

contrasts(data_fw$a) <- apcR::create.poly
contrasts(data_fw$p) <- apcR::create.poly
contrasts(data_fw$c) <- apcR::create.poly
# zero out the period linear effect
contrasts(data_fw$p)[, 1] <- 0

fw <- lm(wordsum ~ a + c + p, data = data_fw)

# get deviations from linear trend and age/cohort linear effects
# (these are the true linear effects when period linear effect = 0)
fw_age <- get_deviations_w_intercept(fw, "^a", contrasts(data_fw$a))
fw_period <- get_deviations_w_intercept(fw, "^p", contrasts(data_fw$p))
fw_cohort <- get_deviations_w_intercept(fw, "^c", contrasts(data_fw$c))
fw_theta1 <- coef(fw)["a.L"]
fw_theta2 <- coef(fw)["c.L"]
```

## Using weighted orthogonal contrasts

```{r}
data_corrected <- as_tibble(d)

contrasts(data_corrected$a) <- contr.poly.weighted(data_corrected$a, width = 5)
contrasts(data_corrected$p) <- contr.poly.weighted(data_corrected$p, width = 5)
contrasts(data_corrected$c) <- contr.poly.weighted(data_corrected$c, width = 5)
# zero out the period linear effect
contrasts(data_corrected$p)[, 1] <- 0

corrected <- lm(wordsum ~ a + c + p, data = data_corrected)

# get deviations from linear trend and age/cohort linear effects
# (these are the true linear effects when period linear effect = 0)
corrected_age <- get_deviations_w_intercept(corrected, "^a", contrasts(data_corrected$a))
corrected_period <- get_deviations_w_intercept(corrected, "^p", contrasts(data_corrected$p))
corrected_cohort <- get_deviations_w_intercept(corrected, "^c", contrasts(data_corrected$c))
corrected_theta1 <- coef(corrected)["a.L"]
corrected_theta2 <- coef(corrected)["c.L"]
```

## Comparison

The model estimates can be seen in this table:

```{r}
screenreg(list(fw, corrected), single.row = TRUE)
```

We can also plot the non-linear effects:

```{r}
to_plot <- bind_rows(
    tibble(effect = c(corrected_age, corrected_period, corrected_cohort),
        year = c(names(corrected_age), names(corrected_period), names(corrected_cohort)),
        apc = c(rep("Age", length(corrected_age)),
            rep("Period", length(corrected_period)),
            rep("Cohort", length(corrected_cohort))),
        which = "This paper"),
    tibble(effect = c(fw_age, fw_period, fw_cohort),
        year = c(names(fw_age), names(fw_period), names(fw_cohort)),
        apc = c(rep("Age", length(fw_age)),
            rep("Period", length(fw_period)),
            rep("Cohort", length(fw_cohort))),
        which = "Fosse & Winship 2019")) %>%
    mutate(which = factor(which, levels = c("This paper", "Fosse & Winship 2019")),
        apc = factor(apc, levels = c("Age", "Period", "Cohort")),
        year = as.numeric(year))

# comparison of non-linear effects
p <- to_plot %>%
    ggplot(aes(x = year, y = effect, group = which, linetype = which)) +
    facet_wrap("apc", scales = "free_x") +
    geom_point() + geom_line() +
    labs(y = "Non-linear effect", group = "", linetype = "") +
    theme_bw() +
    theme(legend.position = "bottom", axis.title.x = element_blank())
p
```

```{r, include = FALSE}
# ggsave("fosse_winship2019.pdf", p, height = 4, width = 7)
```

## 2D-APC Plot

```{r}
two2d_plot <- function(theta1, theta2) {
    sum_thetas = abs(theta2 - theta1)
    limits = c(-sum_thetas, sum_thetas)

    ggplot() +
        geom_abline(aes(intercept = theta1, slope = -1),
            size = 1, color = "#1080BA") +
        geom_hline(yintercept = 0, linetype = 2, color = "#3C4650") +
        geom_hline(yintercept = -(theta2 - theta1), linetype = 2, color = "#3C4650") +
        geom_vline(xintercept = 0, linetype = 2, color = "#3C4650") +
        scale_y_continuous(expression(alpha),
            limits = limits * 1.2,
            sec.axis = sec_axis(~.+(theta2 - theta1), name = expression(gamma))) +
        scale_x_continuous(expression(pi),
            limits = limits * 1.1,
            sec.axis = dup_axis()) +
        theme_bw() +
        theme(text = element_text(size = 10),
            axis.title.y.left = element_text(angle = 0, vjust = 0.5),
            axis.title.y.right = element_text(angle = 0, vjust = 0.5))
}

two2d_plot(fw_theta1, fw_theta2)
two2d_plot(corrected_theta1, corrected_theta2)
```

```{r, include = FALSE}
# ggsave("fosse_winship2019_2dapc.pdf", p, height = 3, width = 4)
```

## Obtaining smoothed estimates

```{r}
devs <- list()
for (i in seq(ncol(contrasts(data_corrected$c)), 3)) {
    contrasts(data_corrected$c)[, i] <- 0
    reduced <- lm(wordsum ~ a + c + p, data = data_corrected)
    devs_i <- get_deviations_w_intercept(reduced, "^c", contrasts(data_corrected$c))
    devs[[i]] <- tibble(degree = i - 1, year = names(devs_i), deviations = devs_i)
}

cohort_by_degree <- bind_rows(devs) %>%
    bind_rows(tibble(degree = ncol(contrasts(data_corrected$c)),
        year = names(corrected_cohort),
        deviations = corrected_cohort)) %>%
    mutate(year = as.numeric(year))

p <- cohort_by_degree %>%
    ggplot(aes(x = as.numeric(year), y = deviations)) +
    facet_wrap("degree") +
    geom_point() + geom_line() +
    theme_bw()
p
```

## Monotonicity constraint for cohort

```{r}
choose_degree <- 5
non_lins <- filter(cohort_by_degree, degree == choose_degree)$deviations
min_slope <- -min(diff(non_lins)) / 5
total_cohort <- non_lins + contrasts(data_corrected$c)[, 1] * min_slope

cohort_df <- filter(cohort_by_degree, degree == choose_degree) %>%
    transmute(effect = deviations,
        year,
        which = paste0("Non-linear only, poly. of degree ", degree)) %>%
    bind_rows(tibble(effect = total_cohort,
        year = as.numeric(names(total_cohort)),
        which = "Under monotonicity constraint"))

p <- cohort_df %>%
    ggplot(aes(x = year, y = effect, linetype = which)) +
    geom_point() + geom_line() +
    theme_bw() +
    theme(text = element_text(size = 12), legend.position = "bottom") +
    labs(x = "Cohort", y = "Effect", linetype = "")
p
```

```{r, include = FALSE}
# ggsave("fosse_winship2019_monotonicity.pdf", p, height = 4, width = 6)
```
